A forest is more than its trees: haplotypes and inferred ARGs

### 2024-11-08 paper/writeup.tex
added new shiny figures

@@

\begin{figure}[!ht]
	\begin{center}
		\includegraphics[width=0.4\linewidth, height=1.2in]{EH-Ex2.pdf}
		\includegraphics[width=0.4\linewidth, height=1.2in]{EH-extended-Ex2.pdf}
	\end{center}
	\caption{\comment{DELETE}}
%	\caption{\comment{Change figure to have the same arrow with 'Extend Haplotypes' in inkscape for unity's sake.}The extend haplotypes method between tree with more unary nodes. Given two neighboring trees, there exists paths $4\to 6\to8\to10$ and $5\to 9\to 10$ in $F_O$, and in $F_I$ we have paths $4\to 7\to 11\to 10$ and $5\to 12\to 10$. Both the out-forest and in-forest both contain paths with the same initial and terminal nodes, $4\to 10$ and $5\to 10$ respectively, we merge both paths and output the tree sequence on the right.}
	\label{fig:complicated_extending_diagram}
\end{figure}

@@

Since the algorithm needs to take multiple passes over the tree sequence
in each direction,
an important practical quesiton for this algorithm is:
how passes do we need to do?
The algorithm is monotone (spans of ancestral nodes only increase),
so it is guaranteed to require only a finite number of passes,
but it is also not hard to construct pathalogical cases that require an arbitrary number of passes.
\figref{fig:num_passes} shows empirical results from simulated tree sequences,
which suggest that only at most five iterations are needed before the algorithm terminates. Indeed, for even larger sequences \ref{tab:edge-counts} shows that 99\% of all changes to an ARG occur after the first iteration with the algorithm completing after four iterations. \comment{ XXX Do we want to add more here? XXX}

\begin{figure}[!ht]
    \caption{
        Number of edges in the tree sequence (vertical axis)
        after each pass of the algorithm (number of passes on the horizontal axis),
        for tree sequences with different numbers of samples (different lines).
        \comment{TODO}
        \label{fig:num_passes}
    }
\end{figure}

%% Find counter example to global minimum.

@@

\comment{Rename Dissimilarity and Similarity}

@@

Concretely, suppose that the two tree sequences have the same set of breakpoints
between trees,
so that $T_1^{(1)}, \ldots, T_N^{(1)}$ are the trees in $\T_1$
and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
\comment{Should we include an addendum that we can do this for tree sequences with different break points its just a little more tedious?}
For a node $n$ and tree $T$
let $S(T, n)$ denote the set of samples that inherit from $n$ in $T$,
and for a pair of nodes $n_1$ and $n_2$ with $n_1$ in $\T_1$ and $n_2$ in $\T_2$,
define
\begin{align*}
    m(n_1, n_2)
    =
    \sum_{k=1}^N (a_k - a_{k-1}) \ind_{S(T^{(1)}_k(n_1) = S(T^{(2)}_k(n_2)} ,
\end{align*}
which is the total span over which the samples below $n_1$ in $\T_1$
matches the samples below $n_2$ in $\T_2$.
Also let
\begin{align*}
    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} ,
\end{align*}
the total span that node $n$ is present in the marginal trees.
The \emph{similarity} between $\T_1$ and $\T_2$ is then defined to be
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
\end{align*}
where the maximum is over all mappings of nodes in $\T_1$ to nodes in $\T_2$.
(Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
and that there may be no nodes in $\T_1$ that are mapped to some nodes in $\T_2$.)
Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
it's \emph{best matching node} in $\T_2$ as
\begin{align*}
    \alpha(n_1) = \argmax_{n_2 \in N_2} \frac{1}{1+|t_{n_1}-t_{n_2}|}\max_{n_2\in N_2} m(n_1, n_2) ,
\end{align*}
so that
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \sum_{n \in N_1} m(n, \alpha(n)) .
\end{align*}
If the best-matching node is unique, we define $\alpha(n_1)$ to be the node in $T_2$
out of those maximizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
(and if \emph{this} is not unique, we pick an arbitrary one) --
however, this potential ambiguity does not affect the definition of $\similarity(\T_1, \T_2)$.
We then define the \emph{dissimilarity} of $\T_1$ from $\T_2$ by
\begin{align*}
    \dis(\T_1, \T_2)
    =
    \frac{\sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n)))}{\sum_{n\in N_1} s(\T_1,n)} ,
\end{align*}
which is the proportion of total span of all nodes in $\T_1$
for which their descendant samples do \emph{not} match those of their best match in $\T_2$.

Other authors \citep[e.g.,]{kelleher2019inferring} have used
an average Robinson-Foulds distance \citep{robinson1981comparison}
as a measure of disagreement between tree sequences.
This could be computed in a very similar way:
instead of $m(n, \alpha(n))$ define
\begin{align*}
    m'(n, \T_2) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{\exists n_2: s(T^{(2)}_k, n_2) = s(T^{(1)}_k, n)} ,
\end{align*}
the total span over which there is \emph{some} node in $\T_2$ whose set of descendant samples
matches $s(T_1,n)$.
Then the average Robinson-Foulds distance (averaged over locations in the genome)
is
\begin{align*}
    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
\end{align*}
\comment{Check this!}\comment{Not true. I believe you can have the problem again of two nodes in TS1 match with node in TS2}
In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
across all trees, but average Robinson-Foulds distance allows a different node to match
on each tree. \comment{I don't think this is true nor do I understand how $m'$ is truely different than $m$. It seems like $m'$ is just the extension of $m$ to the level of tree sequences.}
The other differences are that \emph{average} Robinson-Foulds distance
normalizes by sequence length, and is symmetrized.
Robinson-Foulds distance between two trees
was defined by \citet{robinson1981comparison}
to the minimum number of edge contraction/expansion operations needed to move
from one tree to the other
(which they then show is equal to the number of edges that induce different splits on the labels).
It would be interesting to define a similar metric on tree sequences,
perhaps using the Subgraph-Prune-and-Regraft moves used by Singer \citet{deng2024robust}.

The dissimilarity $\dis(\T_1, \T_2)$ measures disagreement between \emph{topologies},
but not times.
If the tree sequence is dated, we can naturally use the ``best match'' $\alpha$
to also compare times,
for instance by returning the weighted root-mean-squared error on inferred times as
\begin{align*}
    \text{wRMSE}_t(\T_1, \T_2)
    = \sqrt{\frac{
        \sum_{n \in N_1} s(\T_1,n) \left(t^{(1)}_n - t^{(2)}_{\alpha(n)} \right)^2 
    }{
        \sum_{n \in N_1} s(\T_1,n)
    } } .
\end{align*}
The mean is computed weighting by node span, so that a dating error
is more serious for a node with a longer span.
\comment{TODO: change the definition?
In practice, time estimation errors seem to be closer to homoskedastic on a log scale,
so this would probably be more informative if errors are computed on the log sale.}

@@

\begin{figure}[bht]
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/node_spans_wo_trapped_unary_regions_CDF.pdf}
	\caption{
        Node spans of simulated tree sequence (blue), simplified tree sequence (green), and extended tree sequence (red).
        \comment{TODO: nodes should be in the same order in all three lines,
        using the ordering from the true tree sequence.}
        \comment{TODO: somehow communicate per-node amount/proportion of correctly and incorrectly added span}
    }
	\label{fig:node-spans}
\end{figure}

@@

\begin{figure}
	\begin{center}
		% \includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
		% \includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
        \includegraphics[width=\textwidth]{dissimilarity.pdf}
	\end{center}
    \caption{
        Accuracy and sensitivity of extended haplotypes
        across a range of sample sizes, on a sequence of length $5\times10^7$ (top)
        and sequence lengths, with 1,000 \comment{diploid?} samples (bottom).
        For each, a simulated ARG containing unary haplotype spans
        was (i) \emph{simplified}, removing the unary spans,
        and (ii) \emph{inferred}, using \tsinfer{} on genotypes;
        then each of these had its haplotypes \emph{extended}.
        The inferred, then simplified, ARG is also shown.
        \textbf{Left:} Discrepancy to the true ARG,
        as proportion of haplotypes that are not represented in the true ARG
        (equation~\eqref{eq:discrepancy}).
        \textbf{Right:} Agreement to the true ARG,
        as proportion of the true ARG that is represented.
        The largest number of trees in the bottom row corresponds
        to the sequence length in the top row;
        see text for more details.
        \comment{TODO: make "inferred simplified" green (dotted) not red}
        \comment{TODO: label "inferred extended" "inferred simplfieid extended"??}
        \comment{TODO: also add a "inferred extended" line}
    }
    \label{fig:discrepancy}
\end{figure}

\begin{figure}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_counts_5e7.pdf}
    \caption{
        number of edges, 5e7, 1,000 samples
        \label{fig:inferred_edge_counts}
    }
\end{figure}

@@

\begin{figure}[bht]
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_log.pdf}
	\caption{Correct Unary Span for Extend Haplotypes (log)}
	\label{fig:correct-unary-span-log}
\end{figure}

\begin{figure}
%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/avg_edge_count_over_iters_5e7_1000s.pdf}
	\caption{The average number of edges for an extended tree sequence per iteration of the extend haplotypes algorithm.}
	\label{fig:edges-per-iteration}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_geneconversion.pdf}
	\caption{
        Accuracy of \algorithmref{alg:extend}
        on simulated ARGs with sample size $10^2$, length $10^7$, gene conversion rate $10^{-8}$ and gene tract length = $10^3$.
        The same quantities for simulated ARGs without gene conversion
        are shown in Supplementary \figref{fig:accuracy_on_truth_no_gc}.
    \label{fig:accuracy_on_truth}
}
\end{figure}

\begin{figure}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_recomb.pdf}
	\caption{Accuracy of \algorithmref{alg:extend}
        as in \figref{fig:accuracy_on_truth} except that simulations were done
        without gene conversion. \comment{Add to supplement.}
        \label{fig:accuracy_on_truth_no_gc}
    }
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_notlog.pdf}
	\caption{\comment{DELETE}}
%	\caption{Correct Unary Span for Extend Haplotypes}
	\label{fig:correct-unary-span}
\end{figure}

>>>>

\newcommand*{\tabref}[2][]{%
	\hyperref[{#2}]{%
		Table~\ref*{#2}%
		\ifx\\#1\\%
		\else
		\,#1%
		\fi
	}%
}

@@

@@

Since the algorithm needs to take multiple passes over the tree sequence
in each direction,
an important practical question for this algorithm is:
how passes do we need to do?
The algorithm is monotone (spans of ancestral nodes only increase),
so it is guaranteed to require only a finite number of passes,
but it is also not hard to construct pathological cases that require an arbitrary number of passes.
Counting edges for every iteration suggest that only at most five iterations are needed before the algorithm terminates. Indeed, for even larger sequences \tabref{tab:edge-counts} shows that 99\% of all changes to an ARG occur after the first iteration with the algorithm completing after four iterations. 
%% Find counter example to global minimum.

@@

@@

Concretely, our method can be applied to any two tree sequences,
 as in the case of \figref{fig:conceptual_discrepancy},
 where we consider the pair of trees which exist at that break point.
However, suppose that the two tree sequences have the same set of breakpoints
between trees,
so that $T_1^{(1)}, \ldots, T_N^{(1)}$ are the trees in $\T_1$
and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
For a node $n$ and tree $T$
let $S(T, n)$ denote the set of samples that inherit from $n$ in $T$,
and for a pair of nodes $n_1$ and $n_2$ with $n_1$ in $\T_1$ and $n_2$ in $\T_2$,
define
\begin{align*}
    m(n_1, n_2)
    =
    \sum_{k=1}^N (a_k - a_{k-1}) \ind_{S(T^{(1)}_k(n_1) = S(T^{(2)}_k(n_2)} ,
\end{align*}
which is the total span over which the samples below $n_1$ in $\T_1$
matches the samples below $n_2$ in $\T_2$.
Also let
\begin{align*}
    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} ,
\end{align*}
the total span that node $n$ is present in the marginal trees.
The \emph{similarity} between $\T_1$ and $\T_2$ is then defined to be
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
\end{align*}
where the maximum is over all mappings of nodes in $\T_1$ to nodes in $\T_2$.
(Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
and that there may be no nodes in $\T_1$ that are mapped to some nodes in $\T_2$.)
Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
it's \emph{best matching node} in $\T_2$ as
\begin{align*}
    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
\end{align*}
so that
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \sum_{n \in N_1} m(n, \alpha(n)) .
\end{align*}
If the best-matching node is unique, we define $\alpha(n_1)$ to be the node in $T_2$
out of those maximizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
(and if \emph{this} is not unique, we pick an arbitrary one) --
however, this potential ambiguity does not affect the definition of $\similarity(\T_1, \T_2)$.
We then define the \emph{dissimilarity} of $\T_1$ from $\T_2$ by
\begin{align*}
    \dis(\T_1, \T_2)
    =
    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))),
\end{align*}
which is the proportion of total span of all nodes in $\T_1$
for which their descendant samples do \emph{not} match those of their best match in $\T_2$.

Other authors \citep[e.g.,]{kelleher2019inferring} have used
an average Robinson-Foulds distance \citep{robinson1981comparison}
as a measure of disagreement between tree sequences.
This could be computed in a very similar way:
instead of $m(n, \alpha(n))$ define
\begin{align*}
    m'(n, \T_2) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{\exists n_2: s(T^{(2)}_k, n_2) = s(T^{(1)}_k, n)} ,
\end{align*}
the total span over which there is \emph{some} node in $\T_2$ whose set of descendant samples
matches $s(T_1,n)$.
Then the average Robinson-Foulds distance (averaged over locations in the genome)
is
\begin{align*}
    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
\end{align*}
In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
across all trees, but average Robinson-Foulds distance allows a different node to match
on each tree.
The other differences are that \emph{average} Robinson-Foulds distance
normalizes by sequence length, and is symmetrized.
Robinson-Foulds distance between two trees
was defined by \citet{robinson1981comparison}
to the minimum number of edge contraction/expansion operations needed to move
from one tree to the other
(which they then show is equal to the number of edges that induce different splits on the labels).
It would be interesting to define a similar metric on tree sequences,
perhaps using the Subgraph-Prune-and-Regraft moves used by Singer \citet{deng2024robust}.

The dissimilarity $\dis(\T_1, \T_2)$ measures disagreement between \emph{topologies},
but not times.
If the tree sequence is dated, we can naturally use the ``best match'' $\alpha$
to also compare times,
for instance by returning the weighted root-mean-squared error on inferred times as
\begin{align*}
    \text{wRMSE}_t(\T_1, \T_2)
    = \sqrt{\frac{
        \sum_{n \in N_1} s(\T_1,n) \left(\log t^{(1)}_n - \log t^{(2)}_{\alpha(n)} \right)^2 
    }{
        \sum_{n \in N_1} s(\T_1,n)
    } } .
\end{align*}
The mean is computed weighting by node span, so that a dating error
is more serious for a node with a longer span.

@@

\begin{figure}[bht]
	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/figure-4-2.pdf}
	\caption{
        Node spans of simulated tree sequence (blue), simplified tree sequence (green), and extended tree sequence (red).
        \comment{TODO: nodes should be in the same order in all three lines,
        using the ordering from the true tree sequence.}
        \comment{TODO: somehow communicate per-node amount/proportion of correctly and incorrectly added span}
    }
	\label{fig:node-spans}
\end{figure}

\begin{figure}
	\includegraphics{newplots_wo_ee/figure-4.pdf}
\end{figure}

@@

\begin{figure}
	\begin{center}
		% \includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
		% \includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
        \includegraphics[width=\textwidth]{newplots_wo_ee/dissimilarity.pdf}
	\end{center}
    \caption{
        Accuracy and sensitivity of extended haplotypes
        across a range of sample sizes, on a sequence of length $5\times10^7$ (top)
        and sequence lengths, with 1,000 \comment{diploid?} samples (bottom).
        For each, a simulated ARG containing unary haplotype spans
        was (i) \emph{simplified}, removing the unary spans,
        and (ii) \emph{inferred}, using \tsinfer{} on genotypes;
        then each of these had its haplotypes \emph{extended}.
        The inferred, then simplified, ARG is also shown.
        \textbf{Left:} Discrepancy to the true ARG,
        as proportion of haplotypes that are not represented in the true ARG
        (equation~\eqref{eq:discrepancy}).
        \textbf{Right:} Agreement to the true ARG,
        as proportion of the true ARG that is represented.
        The largest number of trees in the bottom row corresponds
        to the sequence length in the top row;
        see text for more details.
        \comment{TODO: make "inferred simplified" green (dotted) not red}
        \comment{TODO: label "inferred extended" "inferred simplfieid extended"??}
        \comment{TODO: also add a "inferred extended" line}
    }
    \label{fig:discrepancy}
\end{figure}

\begin{figure}
	\includegraphics{newplots_wo_ee/figure5.pdf}
\end{figure}

%\begin{figure}
%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_counts_5e7.pdf}
%    \caption{
%        number of edges, 5e7, 1,000 samples
%        \label{fig:inferred_edge_counts}
%    }
%\end{figure}

@@

%
%\begin{figure}[bht]
%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_log.pdf}
%	\caption{Correct Unary Span for Extend Haplotypes (log)}
%	\label{fig:correct-unary-span-log}
%\end{figure}
%
%\begin{figure}
%%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/avg_edge_count_over_iters_5e7_1000s.pdf}
%	\caption{The average number of edges for an extended tree sequence per iteration of the extend haplotypes algorithm.}
%	\label{fig:edges-per-iteration}
%\end{figure}

@@

%\begin{figure}
%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_geneconversion.pdf}
%	\caption{
%        Accuracy of \algorithmref{alg:extend}
%        on simulated ARGs with sample size $10^2$, length $10^7$, gene conversion rate $10^{-8}$ and gene tract length = $10^3$.
%        The same quantities for simulated ARGs without gene conversion
%        are shown in Supplementary \figref{fig:accuracy_on_truth_no_gc}.
%    \label{fig:accuracy_on_truth}
%}
%\end{figure}
%
%\begin{figure}
%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_recomb.pdf}
%	\caption{Accuracy of \algorithmref{alg:extend}
%        as in \figref{fig:accuracy_on_truth} except that simulations were done
%        without gene conversion. \comment{Add to supplement.}
%        \label{fig:accuracy_on_truth_no_gc}
%    }
%\end{figure}

@@
